Investigating Fault-Proneness Through Hotfix Commits in the Light of Different Programming Languages

### 2016-02-20 msr_paper/main.tex
Revised on section 3-5 by adding the method to identify hotfixes

\subsection{Dataset}
The SourceForge13 Dataset is selected for this study. It consists of \textbf{49865} projects, selected from the top 16 most popular project types in the SourceForge dataset. That enables us to differentiate results for the different kinds of projects.

\subsection{Approach for identifying hotfix commits}

@@

\noindent\textbf{RQ2 (hotfixes vs project categories):} What project categories achieve the highest/lowest number of hotfixes per version update?

\noindent\textbf{RQ3 (hotfixes vs languages):} What languages achieve the highest/lowest number of hotfixes per version update?

\noindent\textbf{RQ4 (hotfixes vs language types):} Are there any relationship between the occurrences of hotfixes and the type of the programming language?

@@

		\hline
		Average & &  & & \textbf{0.25}&\textbf{0.17}&\textbf{0.28}&\textbf{0.28}&\textbf{0.28}&\textbf{0.47}&\textbf{0.33}&\textbf{0.17}&\textbf{0.12}&\textbf{0.32}&\textbf{0.27}&\textbf{0.34}&\textbf{0.53}&\textbf{0.11}&\textbf{0.27}&\textbf{0.32} \\
		\hline
	\end{tabular}
	\caption{Ration of hotfix per version update of 8 common languages and their 16 respective domains. $\mathcal{P}$, $\mathcal{C}$, and $\mathcal{T}$ represent the classes of the language. $\mathcal{P}$ stands for Programming Paradigm class (P=Procedural, S=Scripting). $\mathcal{C}$ is for Compilation class (S=Static, D=Dynamic), and $\mathcal{T}$ is for Type class (W=weakly-typed, S=Strongly-typed).}
	\label{tab:1}
\end{table*}

\section{Results}
We performed an empirical study on the SourceForge13 dataset aiming to find the answers for our 4 research questions. The findings of each research question is explained below.
\subsection{RQ1: language popularity}
Our investigation on the most widely used languages across the top 16 most popular categories revealed that there are surprisingly few such languages. We identified only 8 widely used languages: Java, C, C++, Perl, Python, PHP, C\#, and JavaScript. The criteria of choosing the languages is based on whether the language presents in most of the categories. We found that there is a significant drop in the amount of occurrence after JavaScript.  %One would expect JavaScript to appear in the list. However, it is present only in 7 categories and hence is ignored from the list. This may be due to the fact that JavaScript is mainly used only for website development while our 16 popular categories span over various aspects of software.

From Table \ref{tab:1}, \textbf{the most widely used languages from the list is Java which appears in all the 16 categories.} C, and C++ share the second place with occurrence in 15 out of 16 categories. While C++ is not present in Site Management, C is not present in Frameworks. The rest of the languages in the list are present in 10 categories on average. However, it is interesting to note that the majority of these languages are not used in Testing, which contains only Java, C, and C++. Similarity, only a few languages appears in Code Generators and Site Management categories.

\subsection{RQ2: hotfixes vs project categories}
Next, we investigate which categories achieve the highest or lowest numbers of hotfixes per version update. We proceed to analyse why certain categories might have low numbers of hotfixes and vice versa.

In Table \ref{tab:1}, if we look at the average number of hotfixes per category we see that \textbf{the Internet, Security, Communications, and Systems Administration have the lowest average number of hot fixes per update} out of the entire group of categories. The categories with \textbf{the highest average numbers of hotfixes per version update are Code Generators and Role playing} which have 0.53 and 0.47 average hot fixes per update respectively. Both these values are much higher than the next highest (0.34).

With regards to those with lower average hot fixes per update, there is a reasonable explanation. Applications under the categories as Security, Internet, Communications and Systems administration are all information critical, which means they have critical assets that need to be protected in order to maintain customer's privacy and trust. For example, with Communications applications it is vital that security is implemented in such a way that an adversary cannot intercept or obtain messages only intended for the receiver. This means that applications that fall under such categories should be extensively tested before deployment to reduce the chance of failure. This results in less hot fix being needed after deployment.

Another supporting example is the usage of C in Internet and Security categories has an incredibly low hot fix average (0.03 and 0.05 respectively). This could be down to fully comprehensive testing techniques, but could also be attributed to the fact that within those categories, the language has been used for an extensive period and hence there is a lot of experience with a variety of different bugs, meaning they are less likely to occur during development as experienced programmers are less likely to make those particular mistakes in the first place. This could also be the reason the projects written in Perl have a very low hot fix average for all four categories.

Role Playing applications have a high average number of hot fixes compared to other projects that fall under the Games category. It is possible that the Games category is a general category. It is a lot simpler than those that fall under the role playing category which is why bugs might be more prominent as a result of increased complexity of the game. Moreover, usually this kind of game contains huge amount of configurations of characters, and items in the game. It is common to release a few hotfixes after releasing a new version due to fixing these configurations. We can obviously see that all languages in Role Playing category have high average numbers of hot fixes per update.

On the other hand, for Code generators, C and C++, have very high hot fix averages while  PHP and Java are the lowest. One can speculate that PHP and Java are established languages for writing code generators and due to this engender low hot fix averages. C and C++ might share syntactic details and semantic structure that would result in a similar hot fix average between them.

\subsection{RQ3: hotfixes vs languages}
Language comparison based on the average hotfix ratio over all the categories in Table 1 has shown that \textbf{Java has appeared to have the least average ratio of hotfixes per version update (0.176)} followed by JavaScript (0.209) and Perl (0.218). However, all the mentioned languages show marginally better results with 8\% difference comparing to other common languages taken under analysis: PHP(0.282), C++(0.286), C (0.305) and C\#(0.339). \textbf{C\# has the highest average ratio of hotfixes among all the languages.}

\subsection{RQ4: hotfixes vs language types}
Another interesting research question is comparing the type of each programming language by their average ratio of hotfixes. The classification in Table \ref{tab:1} is inspired by \cite{Ray2014}, where languages were compared based on the number of bug fixes in each language and were divided into three main Classes. The first class (Programming Paradigm) has the following categories: Procedural, Functional, Scripting. The second class is Compilation containing Static and Dynamic type. The final one, Type Class, separates languages into Strongly-typed and Weakly-typed. According to \cite{Ray2014}, strongly-typed languages experienced fewer bug fixes than its counterpart, weakly-typed. Static programming languages required fewer corrective commits than dynamic ones. And finally, functional languages encountered fewer bug fixes than procedural ones.

Due to our criteria of selecting the languages, the chosen 8 languages are all either procedural or scripting, but not functional. Most probably this was due to the fact that functional languages are limited to specific use cases and as such are not widely popular across different domains  \cite{Wadler1999}. Consequently, we compare only Procedural and Scripting languages.

\subsubsection*{Programming paradigm: procedural vs scripting}
From Table \ref{tab:1}, there is no clear trend or pattern of the hotfix ratio among Procedural (Java, C, C\#, C++) and Scripting (Perl, Python, PHP) across all the categories. This is possibly because the number of data points (languages) was too low to observe such behaviours and make any firm conclusions. %It was always the case that at least one of the languages would not follow a pattern. For some categories, the comparison would consider only one language from a certain category.

\subsubsection*{Compilation: static vs dynamic}
In this class, static (Java, C, C\#, C++) and dynamic (Perl, Python, PHP) languages led to the same conclusions of no clear different of hotfix ratio observed between the two sets as can be seen from Table \ref{tab:1}. The reason was because the languages in these two categories coincided with the languages in the first two classes (procedural and scripting).

\subsubsection*{Type: strongly-type vs weakly-typed}
The third class compares strongly-typed (Java, C\#, Perl, Python) and weakly-typed (C, C++, PHP) languages. Similarly to the previous two classes, we encountered no clear pattern across the categories. Furthermore, we found the same problem of having comparisons against only one member of a language class. Solely in the Communications category, we were able to detect some pattern, where strongly-typed languages performed slightly better. However, there were only two languages labelled as weak. Therefore, we cannot reach any satisfying conclusions based on this finding.

@@

We believe that there are still aspects to be improved in our study. In particular, more datasets could be used, including Github (provided that projects are categorised into topics), and more reliable and accurate method to identify bump commits. Instead of using data from SourceForge, real commercial projects which have been open-sourced can be investigated to enhance the quality of our data. In such projects, releases should occur more frequently and thus, there is a higher probability of identifying more hotfixes.

>>>>

\subsection{Dataset}
The Boa September 2013 SourceFourge dataset is selected for this study. It consists of 49,865 projects, selected from the top 16 most popular project categories in the dataset. We only concentrate on projects that incorporate versioning (i.e. containing bump commits). %That enables us to differentiate results of each different project category.

\subsection{Approach for identifying hotfix commits}
We invented a method to distinguish a hotfix based on commit messages. The method has 2 parts:

\textbf{1) Detecting version bump commits:} From the observed data, it turned out to be straightforward to extract this information by a regular expression that matches the following strings: ``bump'', ``version'', ``update'' in an appropriate order.

\textbf{2) Detecting hotfix commits:} Due to the similarity between commit messages for normal bug fixes and hotfix commits, it was challenging to distinguish them only based on the message. Therefore, we applied two additional filters: time and range of commits.

\textbf{Time} -- hotfix commits have to be submitted within 1 day from the version bump commit. Thus, a hotfix is assigned the highest priority and has to be submitted rapidly. Furthermore, it is very likely that fixes performed on the first day after a new release will refer to problems occurring before the version increase.

\textbf{Look-ahead range for commits} -- only n numbers of commits should be considered after a version bump. We focused on the Look-ahead parameter and decided to investigate only the first 10 commits as any commit after this point would possibly be related to code added after the release rather than before that. We have performed an initial investigation of 5, 10 and 15 commits and found that 10 achieved the best balance between high number of fix commits and low number of commits that introduce new features.

@@

\noindent\textbf{RQ2 (hotfixes vs. project categories):} What project categories achieve the highest/lowest number of hotfixes per version update?

\noindent\textbf{RQ3 (hotfixes vs. languages):} What languages achieve the highest/lowest number of hotfixes per version update?

\noindent\textbf{RQ4 (hotfixes vs. language types):} Are there any relationship between the occurrences of hotfixes and the type of the programming language?

@@

		\hline
		Average & &  & & \textbf{0.25}&\textbf{0.17}&\textbf{0.28}&\textbf{0.28}&\textbf{0.28}&\textbf{0.47}&\textbf{0.33}&\textbf{0.17}&\textbf{0.12}&\textbf{0.32}&\textbf{0.27}&\textbf{0.34}&\textbf{0.53}&\textbf{0.11}&\textbf{0.27}&\textbf{0.32} \\
		\hline
	\end{tabular}
	\caption{Ratios of hotfix commits per version bump commits of 8 common languages and their 16 respective categories. $\mathcal{P}$, $\mathcal{C}$, and $\mathcal{T}$ represent the classes of the language. $\mathcal{P}$ stands for Programming Paradigm class (P=Procedural, S=Scripting). $\mathcal{C}$ is for Compilation class (S=Static, D=Dynamic), and $\mathcal{T}$ is for Type class (W=weakly-typed, S=Strongly-typed).}
	\label{tab:1}
\end{table*}

\section{Results}
We performed an empirical study on the SourceForge dataset using our method of identifying version bump commits and hotfix commits. The findings are explained below.
\subsection{RQ1: language popularity}
Our investigation on the most widely used languages across the top 16 most popular categories revealed that there are surprisingly few such languages. We chose the languages based on their frequency of occurrence across 16 categories. From Table \ref{tab:1}, the 8 widely used languages include Java (16), C (15), C++ (15), Perl (11), Python (11), PHP (11), C\# (10), and JavaScript (7) respectively.  We found a significant drop in occurrence for the rest of languages so their results are not included here.  %One would expect JavaScript to appear in the list. However, it is present only in 7 categories and hence is ignored from the list. This may be due to the fact that JavaScript is mainly used only for website development while our 16 popular categories span over various aspects of software.

It is interesting that majority of these languages are not used in Testing, which contains only Java, C, and C++. Similarity, only a few languages appears in Code Generators and Site Management categories.

% From Table \ref{tab:1}, \textbf{the most widely used languages from the list is Java which appears in all the 16 categories.} C, and C++ share the second place with occurrence in 15 out of 16 categories. While C++ is not present in Site Management, C is not present in Frameworks. The rest of the languages in the list are present in 10 categories on average. However, it is interesting to note that the majority of these languages are not used in Testing, which contains only Java, C, and C++. Similarity, only a few languages appears in Code Generators and Site Management categories.

\subsection{RQ2: hotfixes vs. project categories}
To be able to compare among languages, we calculated their hotfix ratios. The hotfix ratio of a language $P$ in category $C$ is a ratio of total hotfix commits per total version bump commits of the language in that category. %We proceed to analyse why certain categories might have low numbers of hotfixes and vice versa.
\begin{equation*}
\mathrm{HotfixRatio}_{P,C} = \frac{\sum{\mathrm{HotfixCommit}_{P,C}}}{\sum{\mathrm{VersionBumpCommit}_{P,C}}}
\end{equation*}
In Table \ref{tab:1}, \textbf{Communications have the lowest average hotfix ratios across all categories (0.11)} followed by Internet, Security, and Systems Administration (0.12, 0.17, 0.17) respectively. These categories are information critical. They are expected to be tested extensively before deployment to reduce the chance of failure. This may results in less hot fixes being needed after deployment. An interesting observation is that the average hotfix ratio of C in Internet category is incredibly low (0.03). We believe that is because of there are several new languages designed specifically for Internet applications. Not many new C code are being written in this area which results in a very low hotfix commits. %Security categories  %This could be down to fully comprehensive testing techniques, but could also be attributed to the fact that within those categories, the language has been used for an extensive period and hence there is a lot of experience with a variety of different bugs, meaning they are less likely to occur during development as experienced programmers are less likely to make those particular mistakes in the first place. This could also be the reason the projects written in Perl have a very low hot fix average for all four categories.

On the other hand, the categories with \textbf{the highest average hotfix ratios are Code Generators (0.53)}. C and C++, have very high average hot fix ratios while  PHP and Java are the lowest. One can speculate that PHP and Java are established languages for writing code generators and due to this engender low hot fix averages. C and C++ might share syntactic details and semantic structure that would result in a similar hot fix average between them.

The second place is Role playing category (0.47) which has higher hotfix ratio than its superset, Game category. We believe that this specific kind of game is more complex with vast amount of configurations of characters and items in the game. It is common to release a few hotfixes after a new version release due to adjusting these configurations. %We can obviously see that all languages in Role Playing category have high average numbers of hot fixes per update.

%Applications under the categories as Security, Internet, Communications and Systems administration are all information critical, which means they have critical assets that need to be protected in order to maintain customer's privacy and trust. For example, with Communications applications it is vital that security is implemented in such a way that an adversary cannot intercept or obtain messages only intended for the receiver. This means that applications that fall under such categories should be extensively tested before deployment to reduce the chance of failure. This results in less hot fix being needed after deployment.

%Role Playing applications have a high average number of hot fixes compared to other projects that fall under the Games category. It is possible that the Games category is a general category. It is a lot simpler than those that fall under the role playing category which is why bugs might be more prominent as a result of increased complexity of the game. Moreover, usually this kind of game contains huge amount of configurations of characters, and items in the game. It is common to release a few hotfixes after releasing a new version due to fixing these configurations. We can obviously see that all languages in Role Playing category have high average numbers of hot fixes per update.

%On the other hand, for Code generators, C and C++, have very high hot fix averages while  PHP and Java are the lowest. One can speculate that PHP and Java are established languages for writing code generators and due to this engender low hot fix averages. C and C++ might share syntactic details and semantic structure that would result in a similar hot fix average between them.

\subsection{RQ3: hotfixes vs. languages}
Language comparison based on the average hotfix ratio over all the categories in Table 1 has shown that \textbf{Java has the least average hotfix ratio (0.176)} followed by JavaScript (0.209) and Perl (0.218). However, all the mentioned languages show marginally better results with approximately 8\% difference comparing to other common languages: PHP (0.282), C++ (0.286), C (0.305) and C\#(0.339). \textbf{C\# has the highest average hotfix ratio among all the languages.}

\subsection{RQ4: hotfixes vs. language types}
Another interesting research question is comparing the type of each programming language by their average hotfix ratio. The classification in Table \ref{tab:1} is inspired by \cite{Ray2014}, where languages were compared using hotfix ratios with respect to their three main classes. Programming Paradigm class (noted as $\mathcal{P}$ in Table \ref{tab:1}) has three variations: procedural, functional, and scripting. The Compilation class contains static and dynamic type. The final one, Type class, separates languages into strongly-typed and weakly-typed. According to \cite{Ray2014}, strongly-typed languages experienced fewer bug fixes than its counterpart, weakly-typed. Static programming languages required fewer corrective commits than dynamic ones. And finally, functional languages encountered fewer bug fixes than procedural ones.

Due to our criteria of selecting the languages, the chosen 8 languages are all either procedural or scripting, but not functional. Most probably this was due to the fact that functional languages are limited to specific use cases and as such are not widely popular across different domains  \cite{Wadler1999}. Consequently, we only compare procedural and scripting languages in this class.

\subsubsection*{Programming paradigm: procedural vs. scripting}
From Table \ref{tab:1}, there is no clear trend or pattern of the hotfix ratio among procedural (Java, C, C\#, C++) and scripting (Perl, Python, PHP) across all the categories. This is possibly because the number of data points (languages) were too low to observe such behaviours and make any firm conclusions. %It was always the case that at least one of the languages would not follow a pattern. For some categories, the comparison would consider only one language from a certain category.

\subsubsection*{Compilation: static vs. dynamic}
In this class, static (Java, C, C\#, C++) and dynamic (Perl, Python, PHP) languages led to the same conclusions of no clear different of hotfix ratio observed between the two sets as can be seen from Table \ref{tab:1}. The reason was because the languages in these two categories coincided with the languages in the first two classes (procedural and scripting).

\subsubsection*{Type: strongly-type vs. weakly-typed}
The third class compares strongly-typed (Java, C\#, Perl, Python) and weakly-typed (C, C++, PHP) languages. Similarly to the previous two classes, we encountered no clear pattern across the categories. Furthermore, we found the same problem of having comparisons against only one member of a language class. Solely in the Communications category, we were able to detect some pattern, where strongly-typed languages performed slightly better. However, there were only two languages labelled as weak. Therefore, we cannot reach any satisfying conclusions based on this finding.

@@

We believe that there are still some aspects to be improved in our study. In particular, more datasets could be used, including Github (provided that projects are categorised into topics), and more reliable and accurate method to identify bump commits. Instead of using data from SourceForge, real commercial projects which have been open-sourced can be investigated to enhance the quality of our data. In such projects, releases should occur more frequently and thus, there is a higher probability of identifying more hotfixes.
